---
title: "Prophage-transformation conflict"
author: "Nicholas Croucher"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE)

library(magrittr)
library(tidyverse)
library(kableExtra)
library(igraph)
library(proxy)
library(purrr)
library(ape)
library(Biostrings)
library(pwalign)
library(ggrepel)
library(glmmTMB)
library(broom)
library(emmeans)
library(cowplot)
library(fitdistrplus)

position_dodge <- function(width = NULL, preserve = "total") {
  ggproto(NULL, PositionDodge,
          width = width,
          preserve = rlang::arg_match0(preserve, c("total", "single"))
  )
}

#' @rdname ggplot2-ggproto
#' @format NULL
#' @usage NULL
#' @export
PositionDodge <- ggproto(
  `_class` = "PositionDodge",
  `_inherit` = ggplot2::Position,
  width = NULL,
  preserve = "total",
  setup_params = function(self, data) {
    flipped_aes <- has_flipped_aes(data)
    data <- flip_data(data, flipped_aes)
    if (is.null(data$xmin) &&
        is.null(data$xmax) && is.null(self$width)) {
      cli::cli_warn(c("Width not defined",
                      "i" = "Set with {.code position_dodge(width = ...)}"))
    }
    
    if (identical(self$preserve, "total")) {
      n <- NULL
    } else {
      panels <- unname(split(data, data$PANEL))
      ns <- vapply(panels,
                   function(panel) {
                     # we need to find the maximum number of times that the
                     # same xmin is used, but making sure we count each group
                     # only once
                     if (is.null(panel$group))
                       return(1)
                     groups <-
                       unname(split(panel, panel$group))
                     xmins <-
                       vapply(groups, function(group)
                         group$xmin[1], double(1))
                     max(table(xmins))
                   },
                   double(1))
      n <- max(ns)
    }
    
    list(width = self$width,
         n = n,
         flipped_aes = flipped_aes)
  },
  
  setup_data = function(self, data, params) {
    data <- flip_data(data, params$flipped_aes)
    if (!"x" %in% names(data) &&
        all(c("xmin", "xmax") %in% names(data))) {
      data$x <- (data$xmin + data$xmax) / 2
    }
    flip_data(data, params$flipped_aes)
  },
  
  compute_panel = function(data, params, scales) {
    data <- flip_data(data, params$flipped_aes)
    collided <- ggplot2:::collide(
      data,
      params$width,
      name = "position_dodge",
      strategy = pos_dodge,
      n = params$n,
      check.width = FALSE
    )
    flip_data(collided, params$flipped_aes)
  }
)

# Dodge overlapping interval.
# Assumes that each set has the same horizontal position.
pos_dodge <- function(df, width, n = NULL) {
  unique_count <- vctrs::vec_unique_count(df$group)
  if (is.null(n)) {
    n <- unique_count
  }
  
  if (n == 1)
    return(df)
  
  if (!all(c("xmin", "xmax") %in% names(df))) {
    df$xmin <- df$x
    df$xmax <- df$x
  }
  
  d_width <- max(df$xmax - df$xmin)
  
  # Have a new group index from 1 to number of groups.
  # This might be needed if the group numbers in this set don't include all of 1:n
  groupidx <- match(df$group, sort(ggplot2:::unique0(df$group)))

  # Find the center for each group, then use that to calculate xmin and xmax
  df$x <- df$x + width * ((groupidx - 0.5) / unique_count - .5)
  df$xmin <- df$x - d_width / n / 2
  df$xmax <- df$x + d_width / n / 2
  
  df
}

generate_mst_from_mlst <- function(mlst_df) {
  
  # Build matrix
  profile_mat <-
    as.matrix(mlst_df %>% 
              tibble::column_to_rownames("ST"))

  # Calculate Hamming distances
  profile_distances_mat <-
    as.matrix(proxy::dist(profile_mat,
                          method = function(a, b) sum(a != b)) )
  
  # Indices where distance == 1
  slv_idx <- which(profile_distances_mat == 1, arr.ind = TRUE)
  
  # Keep only upper triangle to avoid duplicate pairs (i<j)
  slv_idx <- slv_idx[slv_idx[, 1] < slv_idx[, 2], , drop = FALSE]
  
  # Build the edge list data frame: two profile name columns + distance
  slv_connections <- data.frame(
    mlst_1 = rownames(profile_distances_mat)[slv_idx[, 1]],
    mlst_2 = colnames(profile_distances_mat)[slv_idx[, 2]]
  )
  
  slv_graph <- graph_from_data_frame(
    d = slv_connections,  # igraph uses 'weight' as the edge attribute name
    directed = FALSE
  )
  
  mlst_mst <- igraph::mst(slv_graph)
  return(mlst_mst)
  
}

plot_mst <- function(mlst_mst) {
  
  ggraph::ggraph(mlst_mst, layout = "stress") +
    ggraph::geom_edge_link(color = "grey70") +
    ggraph::geom_node_point(size = 1.5, color = "grey30") +
    theme_void()

}

# Normalize bases to uppercase and map U->T, "."->"-"
normalise_bases <- function(x) {
  x <- toupper(x)
  x[x == "U"] <- "T"
  x[x == "."] <- "-"
  x
}

# Allowed IUPAC DNA bases + gap
allowed <- c("A","C","G","T","N","R","Y","S","W","K","M","B","D","H","V","-")

# Validate a per-base character vector
is_valid_base_vec <- function(x) {
  x2 <- normalise_bases(x)
  all(x2 %in% allowed)
}

# Convert DNAbin or per-base character vector to a sanitized DNA string; return NULL if invalid
safe_dna_string_or_null <- function(x) {
  s <- if (inherits(x, "DNAbin")) {
    paste0(as.character(x), collapse = "")
  } else if (is.character(x)) {
    paste0(x, collapse = "")
  } else {
    return(NULL)
  }
  s <- toupper(s)
  s <- chartr("U", "T", s)
  s <- gsub("\\.", "-", s)
  invalid <- gsub("[ACGTNRYSWKMBVDH-]", "", s)
  if (nchar(invalid) > 0) return(NULL)
  return(s)
}

# Align two sequences; return a tibble of mismatches or NULL if invalid
mismatch_tbl_aligned <- function(seq_from, seq_to, ignore = c("-", "N")) {
  s_from <- safe_dna_string_or_null(seq_from)
  s_to   <- safe_dna_string_or_null(seq_to)
  if (is.null(s_from) || is.null(s_to)) return(NULL)

  aln <- pairwiseAlignment(DNAString(s_from), DNAString(s_to),
                           type = "global", substitutionMatrix = NULL,
                           gapOpening = 5, gapExtension = 2)

  a_from <- as.character(pattern(aln))
  a_to   <- as.character(subject(aln))
  cf <- normalise_bases(strsplit(a_from, "")[[1]])
  ct <- normalise_bases(strsplit(a_to, "")[[1]])

  keep <- !(cf %in% ignore | ct %in% ignore)
  cf2 <- cf[keep]; ct2 <- ct[keep]
  pos <- which(cf2 != ct2)

  if (length(pos) == 0) {
    # Return an empty tibble (this is a valid comparison with zero mismatches)
    return(tibble(pos = integer(), base_from = character(), base_to = character()))
  }

  return(tibble(pos = pos, base_from = cf2[pos], base_to = ct2[pos]))
}

# Compare two sequences; return tibble of mismatches or NULL to skip invalid pairs
mismatch_tbl_dnabin <- function(seq_from, seq_to, ignore = c("-", "N")) {
  if (is.null(seq_from) || is.null(seq_to)) return(NULL)

  cf <- normalise_bases(as.character(seq_from))
  ct <- normalise_bases(as.character(seq_to))

  # Skip invalid characters by returning NULL
  if (!is_valid_base_vec(cf) || !is_valid_base_vec(ct)) return(NULL)

  # Align if lengths differ; otherwise compare directly
  #if (length(cf) != length(ct)) {
    return(mismatch_tbl_aligned(seq_from, seq_to, ignore = ignore))
  #}

  keep <- !(cf %in% ignore | ct %in% ignore)
  cf2 <- cf[keep]; ct2 <- ct[keep]
  pos <- which(cf2 != ct2)

  if (length(pos) == 0) {
    return(tibble(pos = integer(), base_from = character(), base_to = character()))
  }

  return(tibble(pos = pos, base_from = cf2[pos], base_to = ct2[pos]))
}

read_with_seqinr_dnabin_list <- function(files) {
  out <- structure(list(), class = c("DNAbin","list"))
  for (f in files) {
    x <- seqinr::read.fasta(f, as.string = FALSE, seqtype = "DNA", set.attributes = FALSE)
    for (nm in names(x)) out[[nm]] <- ape::as.DNAbin(x[[nm]])
  }
  return(out)
}

identify_mismatches_between_slvs <- function(mlst_profile_df, slvs_df, fn = NULL) {

  # Identify the allele (locus) columns
  allele_cols <- setdiff(names(mlst_profile_df), "ST")
  
  # Identify the allele changes between SLVs
  allele_changes_df <-
    slvs_df %>%
      # Avoid name clash between slvs$from/to and allele value columns named "from"/"to"
      dplyr::rename(from_ST = from, to_ST = to) %>%
      # Join profiles for the "from" ST
      dplyr::inner_join(
        mlst_profile_df %>% 
          dplyr::mutate(ST = as.character(ST)) %>% 
          dplyr::rename_with(~ paste0(., "_from"), all_of(allele_cols)),
        by = c("from_ST" = "ST")
      ) %>%
      # Join profiles for the "to" ST
      dplyr::inner_join(
        mlst_profile_df %>% 
          dplyr::mutate(ST = as.character(ST)) %>% 
          dplyr::rename_with(~ paste0(., "_to"), all_of(allele_cols)),
        by = c("to_ST" = "ST")
      ) %>%
      # Reshape so that each row becomes multiple rows, one per locus,
      # with two columns: `from` (allele number) and `to` (allele number)
      tidyr::pivot_longer(
        cols = all_of(c(paste0(allele_cols, "_from"), paste0(allele_cols, "_to"))),
        names_to = c("locus", ".value"),
        names_pattern = "^(.*)_(from|to)$"
      ) %>%
      # Keep the single differing locus (for true SLV edges)
      dplyr::filter(!is.na(from) & !is.na(to) & from != to) %>%
      # Present clean output
      dplyr::select(from_ST, to_ST, locus, allele_from = from, allele_to = to)
  
  mlst_sequences_dna <-
    read_with_seqinr_dnabin_list(
      Sys.glob(fn)
    )
  
  mismatches_df <-
    allele_changes_df %>%
      dplyr::mutate(
        seq_from_id = paste0(locus, "_", allele_from),
        seq_to_id   = paste0(locus, "_", allele_to),
        mismatches  = purrr::pmap(
          list(seq_from_id, seq_to_id),
          function(id_from, id_to) {
            mismatch_tbl_dnabin(mlst_sequences_dna[[id_from]], mlst_sequences_dna[[id_to]])
          }
        ),
        is_invalid  = purrr::map_lgl(mismatches, is.null),
        n_mismatch  = map_int(mismatches,
                              ~ if (is.null(.x)) NA_integer_ else nrow(.x))
      ) %>% 
    dplyr::mutate(
      Classification =
        dplyr::case_when(
          n_mismatch > 2 ~ "Recombination",
          n_mismatch > 0 ~ "Point mutation",
          n_mismatch == 0 ~ "Different lengths"
        )
    ) %>% 
    dplyr::filter(Classification != "Different lengths")
  
}

```

# Analyse genotyping data

The species to be analysed are taken from those found to contribute to the burden of [antimicrobial resistant-disease](https://doi.org/10.1016/S0140-6736(21)02724-0), as these constitute a set of genetically-diverse and well-sampled bacteria:

- *Escherichia coli*: Multiple MLST schemes available for [*Escherichia* spp.](https://pubmlst.org/bigsdb?db=pubmlst_escherichia_seqdef); the Achtman scheme was used, as it had 7 loci.

- *Staphylococcus aureus*: One MLST database available for [*S. aureus*](https://pubmlst.org/bigsdb?db=pubmlst_saureus_seqdef)

- *Klebsiella pneumoniae*: One MLST database available for [*K. pneumoniae*](https://bigsdb.pasteur.fr/cgi-bin/bigsdb/bigsdb.pl?db=pubmlst_klebsiella_seqdef)

- *Streptococcus pneumoniae*: One MLST database available for [*S. pneumoniae*](https://pubmlst.org/bigsdb?db=pubmlst_spneumoniae_seqdef)

- *Acinetobacter baumanii*: One MLST database available for [*A. baumanii*](https://pubmlst.org/bigsdb?db=pubmlst_abaumannii_seqdef)

- *Pseudomonas aeruginosa*: One MLST database available for [*P. aeruginosa*](https://pubmlst.org/bigsdb?db=pubmlst_paeruginosa_seqdef)

- *Mycobacterium tuberculosis*: Excluded, as low diversity means MLST is not informative for this species.

- *Enterococcus faecium*: One MLST database available for [*E. faecium*](https://pubmlst.org/bigsdb?db=pubmlst_efaecium_seqdef), although it is not clear whether other enterococci are effectively excluded, based on the species label in the profiles.

- *Enterobacter* spp.: One MLST database available for [*E. cloacae* and related species](https://pubmlst.org/bigsdb?db=pubmlst_ecloacae_seqdef).

- *Streptococcus agalactiae*: One MLST database available for [*S. agalactiae*](https://pubmlst.org/bigsdb?db=pubmlst_sagalactiae_seqdef).

- *Salmonella spp.*: Dominated by *S. enterica*. One MLST database available for [*S. enterica*](https://pubmlst.org/bigsdb?db=pubmlst_salmonella_seqdef).

- *Enterococcus faecalis*: One MLST database available for [*E. faecalis*](https://pubmlst.org/bigsdb?db=pubmlst_efaecalis_seqdef).

- *Proteus* spp.: One MLST database for [*Proteus* species](https://pubmlst.org/bigsdb?db=pubmlst_proteus_seqdef).

- *Serratia* spp.: One MLST database for [*Serratia* species](https://pubmlst.org/bigsdb?db=pubmlst_serratia_seqdef).

- *Streptococcus pyogenes*: One MLST database available for [*S. pyogenes*](https://pubmlst.org/bigsdb?db=pubmlst_spyogenes_seqdef).

- *Citrobacter* spp.: One MLST database available for [*C. freundii* and related species](https://pubmlst.org/bigsdb?db=pubmlst_cfreundii_seqdef).

- *Haemophilus influenzae*: One MLST database available for [*H. influenzae*](https://pubmlst.org/bigsdb?db=pubmlst_hinfluenzae_seqdef).

- *Shigella* spp.: Excluded as not evolutionary independent from *E. coli*.

- *Morganella morganii*: One MLST database available for [*M. morganii*](https://bigsdb.pasteur.fr/cgi-bin/bigsdb/bigsdb.pl?db=pubmlst_morganella_seqdef).

These were supplemented with other species representing genera [known to be naturally transformable](https://doi.org/10.1016/j.resmic.2007.09.004):

- *Agrobacterium tumefaciens*:  Excluded, as no MLST data available.

- *Neisseria* spp.: One MLST database available for [*Neisseria* species](https://pubmlst.org/bigsdb?db=pubmlst_neisseria_seqdef); this can be included, as both *N. meningitidis* and *N. gonorrhoeae* are naturally transformable.

- *Legionella pneumophila*: Excluded, as no MLST data available from BigsDB.

- *Moraxella* spp.: One MLST database available for [*M. catarrhalis*](https://pubmlst.org/bigsdb?db=pubmlst_moraxella_seqdef); this can be included, as *Moraxella* species are generally regarded as transformable.

- *Vibrio cholerae*: One MLST database available for [*V. cholerae*](https://pubmlst.org/bigsdb?db=pubmlst_vcholerae_seqdef).

- *Campylobacter jejuni* & *coli*: One MLST database available for [*Campylobacter* spp.](https://pubmlst.org/bigsdb?db=pubmlst_campylobacter_seqdef). Both species are transformable, and therefore can be considered jointly.

- *Helicobacter pylori*: One MLST database available for [*H. pylori*](https://pubmlst.org/bigsdb?db=pubmlst_helicobacter_seqdef).

- *Mycobacterium smegmatis*: Excluded as no MLST scheme available.

- *Bacillus subtilis*: One MLST database available for [*B. subtilis*](https://pubmlst.org/bigsdb?db=pubmlst_bsubtilis_seqdef).

Searches run on 4/12/25.

``` {r Define links between datasets}

# Data structures
species_name <- list()
display_name <- list()
st_prefix <- list()
lower_length_threshold <- list()
upper_length_threshold <- list()
transformable <- list()
primary_niche <- list()

# E. coli
species_name[["Eco"]] = "Escherichia coli"
display_name[["Eco"]] = "italic('E. coli')"
st_prefix[["Eco"]] = "Eco_Achtman"
lower_length_threshold[["Eco"]] = 4000000
upper_length_threshold[["Eco"]] = 6500000
transformable[["Eco"]] = "Non-transformable"
primary_niche[["Eco"]] = "Gastrointestinal tract"
# Citrobacter species
species_name[["Cit"]] = "Citrobacter spp"
display_name[["Cit"]] = "italic('Citrobacter')~'spp.'"
st_prefix[["Cit"]] = "Cit"
lower_length_threshold[["Cit"]] = 4000000
upper_length_threshold[["Cit"]] = 6500000
transformable[["Cit"]] = "Non-transformable"
primary_niche[["Cit"]] = "Gastrointestinal tract"
# Enterobacter species
species_name[["Esp"]] = "Enterobacter spp"
display_name[["Esp"]] = "italic('Enterobacter')~'spp.'"
st_prefix[["Esp"]] = "Esp"
lower_length_threshold[["Esp"]] = 4000000
upper_length_threshold[["Esp"]] = 6500000
transformable[["Esp"]] = "Non-transformable"
primary_niche[["Esp"]] = "Gastrointestinal tract"
# Proteus species
species_name[["Psp"]] = "Proteus spp"
display_name[["Psp"]] = "italic('Proteus')~'spp.'"
st_prefix[["Psp"]] = "Psp"
lower_length_threshold[["Psp"]] = 3000000
upper_length_threshold[["Psp"]] = 5000000
transformable[["Psp"]] = "Non-transformable"
primary_niche[["Psp"]] = "Gastrointestinal tract"
# Serratia species
species_name[["Ssp"]] = "Serratia spp"
display_name[["Ssp"]] = "italic('Serratia')~'spp.'"
st_prefix[["Ssp"]] = "Ssp"
lower_length_threshold[["Ssp"]] = 4000000
upper_length_threshold[["Ssp"]] = 6500000
transformable[["Ssp"]] = "Non-transformable"
primary_niche[["Ssp"]] = "Gastrointestinal tract"
# S. aureus
species_name[["Sau"]] = "Staphylococcus aureus"
display_name[["Sau"]] = "italic('S. aureus')"
st_prefix[["Sau"]] = "Sau"
lower_length_threshold[["Sau"]] = 2000000
upper_length_threshold[["Sau"]] = 4000000
transformable[["Sau"]] = "Non-transformable"
primary_niche[["Sau"]] = "Multiple"
# K. pneumoniae
species_name[["Kpn"]] = "Klebsiella pneumoniae"
display_name[["Kpn"]] = "italic('K. pneumoniae')"
st_prefix[["Kpn"]] = "Kpn"
lower_length_threshold[["Kpn"]] = 4000000
upper_length_threshold[["Kpn"]] = 7000000
transformable[["Kpn"]] = "Non-transformable"
primary_niche[["Kpn"]] = "Gastrointestinal tract"
# S. pneumoniae
species_name[["Spn"]] = "Streptococcus pneumoniae"
display_name[["Spn"]] = "italic('S. pneumoniae')"
st_prefix[["Spn"]] = "Spn"
lower_length_threshold[["Spn"]] = 1500000
upper_length_threshold[["Spn"]] = 3000000
transformable[["Spn"]] = "Transformable"
primary_niche[["Spn"]] = "Respiratory tract"
# A. baumanii
species_name[["Abau"]] = "Acinetobacter baumanii"
display_name[["Abau"]] = "italic('A. baumanii')"
st_prefix[["Abau"]] = "Abau_Oxford"
lower_length_threshold[["Abau"]] = 3000000
upper_length_threshold[["Abau"]] = 5500000
transformable[["Abau"]] = "Transformable"
primary_niche[["Abau"]] = "Environmental"
# P. aeruginosa
species_name[["Pae"]] = "Pseudomonas aeruginosa"
display_name[["Pae"]] = "italic('P. aeruginosa')"
st_prefix[["Pae"]] = "Pae"
lower_length_threshold[["Pae"]] = 5500000
upper_length_threshold[["Pae"]] = 8500000
transformable[["Pae"]] = "Transformable"
primary_niche[["Pae"]] = "Environmental"
# E. faecium
species_name[["Efm"]] = "Enterococcus faecium"
display_name[["Efm"]] = "italic('E. faecium')"
st_prefix[["Efm"]] = "Efm"
lower_length_threshold[["Efm"]] = 1500000
upper_length_threshold[["Efm"]] = 4000000
transformable[["Efm"]] = "Non-transformable"
primary_niche[["Efm"]] = "Gastrointestinal tract"
# S. agalactiae
species_name[["Sag"]] = "Streptococcus agalactiae"
display_name[["Sag"]] = "italic('S. agalactiae')"
st_prefix[["Sag"]] = "Sag"
lower_length_threshold[["Sag"]] = 1500000
upper_length_threshold[["Sag"]] = 3000000
transformable[["Sag"]] = "Non-transformable"
primary_niche[["Sag"]] = "Multiple"
# S. enterica
species_name[["Sen"]] = "Salmonella enterica"
display_name[["Sen"]] = "italic('S. enterica')"
st_prefix[["Sen"]] = "Sen"
lower_length_threshold[["Sen"]] = 3500000
upper_length_threshold[["Sen"]] = 6000000
transformable[["Sen"]] = "Non-transformable"
primary_niche[["Sen"]] = "Gastrointestinal tract"
# E. faecalis
species_name[["Efs"]] = "Enterococcus faecalis"
display_name[["Efs"]] = "italic('E. faecalis')"
st_prefix[["Efs"]] = "Efs" # not run
lower_length_threshold[["Efs"]] = 2000000
upper_length_threshold[["Efs"]] = 4000000
transformable[["Efs"]] = "Non-transformable"
primary_niche[["Efs"]] = "Gastrointestinal tract"
# S. pyogenes
species_name[["Spy"]] = "Streptococcus pyogenes"
display_name[["Spy"]] = "italic('S. pyogenes')"
st_prefix[["Spy"]] = "Spy"
lower_length_threshold[["Spy"]] = 1000000
upper_length_threshold[["Spy"]] = 2500000
transformable[["Spy"]] = "Non-transformable"
primary_niche[["Spy"]] = "Respiratory tract"
# H. influenzae
species_name[["Hflu"]] = "Haemophilus influenzae"
display_name[["Hflu"]] = "italic('H. influenzae')"
st_prefix[["Hflu"]] = "Hflu"
lower_length_threshold[["Hflu"]] = 1000000
upper_length_threshold[["Hflu"]] = 2500000
transformable[["Hflu"]] = "Transformable"
primary_niche[["Hflu"]] = "Respiratory tract"
# M. morganii
species_name[["Mmo"]] = "Morganella morganii"
display_name[["Mmo"]] = "italic('M. morganii')"
st_prefix[["Mmo"]] = "Mmo"
lower_length_threshold[["Mmo"]] = 3000000
upper_length_threshold[["Mmo"]] = 5000000
transformable[["Mmo"]] = "Non-transformable"
primary_niche[["Mmo"]] = "Gastrointestinal tract"
# Neisseria species
species_name[["Nsp"]] = "Neisseria spp"
display_name[["Nsp"]] = "italic('Neisseria')~'spp.'"
st_prefix[["Nsp"]] = "Nsp"
lower_length_threshold[["Nsp"]] = 1500000
upper_length_threshold[["Nsp"]] = 3000000
transformable[["Nsp"]] = "Transformable"
primary_niche[["Nsp"]] = "Respiratory tract"
# Moraxella species
species_name[["Msp"]] = "Moraxella catarrhalis"
display_name[["Msp"]] = "italic('Moraxella')~'spp.'"
st_prefix[["Msp"]] = "Msp"
lower_length_threshold[["Msp"]] = 1000000
upper_length_threshold[["Msp"]] = 3000000
transformable[["Msp"]] = "Transformable"
primary_niche[["Msp"]] = "Respiratory tract"
# Campylobacter species
species_name[["Cam"]] = "Campylobacter spp"
display_name[["Cam"]] = "italic('Campylobacter')~'spp.'"
st_prefix[["Cam"]] = "Cam"
lower_length_threshold[["Cam"]] = 1000000
upper_length_threshold[["Cam"]] = 3000000
transformable[["Cam"]] = "Transformable"
primary_niche[["Cam"]] = "Gastrointestinal tract"
# Helicobacter pylori
species_name[["Hpy"]] = "Helicobacter pylori"
display_name[["Hpy"]] = "italic('H. pylori')"
st_prefix[["Hpy"]] = "Hpy"
lower_length_threshold[["Hpy"]] = 1000000
upper_length_threshold[["Hpy"]] = 2500000
transformable[["Hpy"]] = "Transformable"
primary_niche[["Hpy"]] = "Gastrointestinal tract"
# Vibrio cholerae
species_name[["Vch"]] = "Vibrio cholerae"
display_name[["Vch"]] = "italic('V. cholerae')"
st_prefix[["Vch"]] = "Vch"
lower_length_threshold[["Vch"]] = 3000000
upper_length_threshold[["Vch"]] = 5000000
transformable[["Vch"]] = "Transformable"
primary_niche[["Vch"]] = "Environmental"
# Bacillus subtilis
species_name[["Bsu"]] = "Bacillus subtilis"
display_name[["Bsu"]] = "italic('B. subtilis')"
st_prefix[["Bsu"]] = "Bsu"
lower_length_threshold[["Bsu"]] = 3000000
upper_length_threshold[["Bsu"]] = 5500000
transformable[["Bsu"]] = "Transformable"
primary_niche[["Bsu"]] = "Environmental"

```

## Calculate MSTs from MLST data

```{r Calculate MST, eval = FALSE}

mlst_profiles <- list()
mlst_msts <- list()
slvs <- list()
mst_plots <- list()

for (name in rev(names(st_prefix))) {
  
  message(paste("Processing",name))
  
  mlst_profiles[[name]] <-
    read.table(paste0("./MLST_data/",st_prefix[[name]],".tab.txt"),
               sep = "\t", header = TRUE) %>% 
      dplyr::select(-dplyr::any_of(c("clonal_complex","species")))
  
  mlst_msts[[name]] <-
    generate_mst_from_mlst(mlst_profiles[[name]])
  
  slvs[[name]] <- as_data_frame(mlst_msts[[name]], what = "edges")
  
  mst_plots[[name]] <- plot_mst(mlst_msts[[name]])
  
  ggsave(paste0("./ebursts/",name,"_eburst.png"),
       mst_plots[[name]],
       width = 8,
       height = 8)
  
}

# Get summaries of MLST data
st_counts_df <-
  enframe(mlst_profiles) %>%
    dplyr::mutate(`Sequence types` =
                    map_int(value, nrow)) %>%  # count rows in each df
    dplyr::select(name, `Sequence types`) %>% 
    dplyr::left_join(enframe(species_name) %>% 
                       tidyr::unnest(value),
                     by = "name") %>% 
    dplyr::mutate(name = value) %>% 
    dplyr::select(-value)

write_csv(
  st_counts_df,
  file = "./MLST_data/st_counts.csv"
)

```

``` {r Calculate mismatches, eval = FALSE}

mismatches <- list()

for (name in rev(names(st_prefix))) {
  
  mismatches[[name]] <-
    identify_mismatches_between_slvs(mlst_profiles[[name]],
                                     slvs[[name]],
                                     fn = paste0("./MLST_sequences/",st_prefix[[name]],"_*.fas")
    )
  
}

mismatches_df <-
  dplyr::bind_rows(mismatches,.id = "Species")

write_csv(
  mismatches_df,
  file = "./MLST_data/mismatches.csv"
)

write_csv(
  mismatches_df %>%
    dplyr::select(Species,seq_from_id,seq_to_id,mismatches) %>% 
    dplyr::distinct() %>% 
    tidyr::unnest(mismatches),
  file = "./MLST_data/mismatch_details.csv"
)

```

``` {r Load mismatches and plot, fig.width = 8, fig.height = 8}

mismatches_df <-
  read_csv("./MLST_data/mismatches.csv") %>% 
    dplyr::select(-mismatches) %>% 
    dplyr::left_join(
      read_csv("./MLST_data/mismatch_details.csv") %>% 
        tidyr::nest(data = c("pos","base_from","base_to")) %>% 
        dplyr::rename(mismatches = data),
      by = c("Species","seq_from_id","seq_to_id")
    ) %>%
        dplyr::left_join(enframe(display_name) %>%
                           tidyr::unnest(value) %>% 
                           dplyr::rename(display = value),
                             by = c("Species" = "name"))

ggplot(mismatches_df,
       aes(x = n_mismatch)) +
  geom_histogram(colour = "blue",fill = "blue") +
  scale_x_continuous(trans = "log10") +
  geom_vline(xintercept = 2.5, col = 'red', linetype = 2) +
  facet_wrap(~display,
             scales = 'free_y',
             labeller = label_parsed) +
  ylab("Count") +
  xlab("Number of mismatches") +
  theme_bw() +
  theme(strip.background = element_blank())

```

``` {r Graph recombination}

rho_m_df <-
  mismatches_df %>% 
        dplyr::filter(!is.na(Classification)) %>% 
        dplyr::group_by(Species) %>% 
        dplyr::summarise(rho_m = sum(Classification=="Recombination")/sum(Classification=="Point mutation")) %>%
        dplyr::ungroup()

species_levels <-
  rho_m_df %>% 
    dplyr::arrange(rho_m) %>% 
    dplyr::pull(Species)

rho_m_df %<>%
  dplyr::mutate(Species = factor(Species, levels = species_levels))

ggplot(rho_m_df,
        aes(x = Species,
            y = rho_m)) +
  geom_col() +
  theme_bw()

```

``` {r Model the probability of changes occurring through recombination, fig.width = 8}

rec_modelling_df <-
  mismatches_df %>% 
    dplyr::filter(!is_invalid) %>% 
    dplyr::select(Species,locus,Classification) %>% 
    dplyr::mutate(Locus = paste0(Species,"_",locus)) %>% 
    dplyr::select(-locus) %>% 
    dplyr::group_by(Species) %>% 
    dplyr::mutate(n_loci = n_distinct(Locus)) %>% 
    dplyr::ungroup() %>% 
    dplyr::mutate(FPR = (1/n_loci)**2) %>% 
    dplyr::group_by(Species,Locus,FPR) %>% 
    dplyr::summarise(Count = n(),
                     Rec_count = sum(Classification == "Recombination")) %>% 
    dplyr::ungroup() %>% 
    dplyr::mutate(`Recombination proportion` = Rec_count/Count)

rec_probability_model <-
  glmmTMB::glmmTMB(
    `Recombination proportion` ~ Species + (1|Locus),
    data = rec_modelling_df,
    family = betabinomial(link = "logit"),
    ziformula = ~0,
    dispformula = ~1,
    weights = Count
  )

performance::check_model(rec_probability_model)

```

``` {r Plot probability of recombination, fig.width=8, fig.height = 8}

# Estimated marginal means per species on the response scale (probabilities)
rec_prob_df <-
  emmeans(
    rec_probability_model,
    specs  = ~ Species,
    type   = "response"  # back-transform from logit to probability
  ) %>%
  tibble::as_tibble()

combined_rec_output_df <-
  rec_prob_df %>% 
    dplyr::left_join(rec_modelling_df, by = "Species") %>% 
    dplyr::left_join(rho_m_df, by = "Species")

ggplot(combined_rec_output_df,
       aes(x = rho_m,
           y = prob)) +
  geom_point() +
  ggrepel::geom_label_repel(aes(label = Species)) +
  theme_bw()

```

# Prophage data

## Search for PHASTER data

Searches run on 8/12/25. For multispecies datasets, the species were selected as the binomial names that were more common than the count of isolates within the genus with an undefined species name.

``` {bash, eval = FALSE}

cd prophage
conda activate edirect
for SPECIES in "Escherichia coli" "Staphylococcus aureus" "Klebsiella pneumoniae" "Streptococcus pneumoniae" "Acinetobacter baumanii"   "Pseudomonas aeruginosa" "Enterococcus faecium"    "Streptococcus agalactiae" "Salmonella enterica" "Enterococcus faecalis"    "Streptococcus pyogenes"   "Haemophilus influenzae" "Morganella morganii"      "Neisseria meningitidis" "Neisseria gonorrhoeae" "Moraxella catarrhalis" "Campylobacter jejuni"     "Helicobacter pylori"      "Vibrio cholerae" "Bacillus subtilis" "Serratia sarumanii" "Serratia ureilytica" "Serratia nevei" "Serratia nematodiphila" "Serratia bockelmannii" "Serratia liquefaciens" "Serratia fonticola" "Serratia marcescens" "Serratia plymuthica" "Proteus mirabilis" "Proteus terrae" "Enterobacter hormaechei" "Citrobacter freundii";
  do echo "Processing $SPECIES";
  SEARCH_STRING=`echo "\"$SPECIES\"[Organism] AND \"latest refseq\"[filter] AND complete_genome[filter] AND (all[filter] NOT anomalous[filter])"`
  esearch -db assembly -query "$SEARCH_STRING" | elink -target nuccore -name assembly_nuccore_refseq | esummary | xtract -pattern DocumentSummary -element AccessionVersion | sort -u > ${SPECIES}_nuccore_complete.txt;
  ./get_prophage_data.sh ${SPECIES}_nuccore_complete.txt ${SPECIES};
done
# Combine multispecies datasets
cat Neisseria\ gonorrhoeae_nuccore_complete.txt Neisseria\ meningitidis_nuccore_complete.txt > ./prophage/Neisseria\ spp_nuccore_complete.txt
cat Campylobacter\ coli_nuccore_complete.txt Campylobacter\ jejuni_nuccore_complete.txt > Campylobacter\ spp_nuccore_complete.txt
cat Serratia*_nuccore_complete.txt > Serratia\ spp_nuccore_complete.txt
cat Proteus*_nuccore_complete.txt > Proteus\ spp_nuccore_complete.txt
cat Enterobacter*_nuccore_complete.txt > Enterobacter\ spp_nuccore_complete.txt
cat Citrobacter*_nuccore_complete.txt > Citrobacter\ spp_nuccore_complete.txt

```

``` {r Parse complete sequences}

phaster_df <-
  read_csv("./phaster_outputs/phaster_precomputed_accessions.csv")

complete_sequences_df <-
  map_dfr(species_name, ~ {
  # Create filename
  filename <- paste0("./prophage/",str_c(str_replace_all(.x, " ", "\\ "), "_nuccore_complete.txt"))
  # Read file if it exists
  if (file.exists(filename)) {
    read_tsv(filename, col_names = "data", col_types = "c") %>%
      mutate(species =.x)
  } else {
    tibble(data = character(0), species =.x)
  }
}) %>% 
  dplyr::filter(
      # Include only complete genomes (shorter accession numbers)
      str_detect(data, "^[A-Z]{2}\\d{4,8}\\.\\d+$")
  ) %>%
  dplyr::mutate(
    # Extract components from data column
    accession_base = str_extract(data, "^[A-Z]{2}\\d+"),
    full_accession = data
  )

phaster_complete_sequences_df <-
  complete_sequences_df %>% 
    # Try exact match first
    dplyr::left_join(
      phaster_df %>%
        dplyr::select(accession, submission_url:summary_url),
      by = c("full_accession" = "accession")
    ) %>%
    # For unmatched, try base accession match
    dplyr::left_join(
      phaster_df %>% 
        dplyr::mutate(accession_base = str_extract(accession, "^[A-Z]{2}\\d+")) %>%
        dplyr::select(accession_base, accession, submission_url:summary_url) %>%
        dplyr::distinct(accession_base,.keep_all = TRUE),
      by = "accession_base",
      suffix = c("", "_base")
    ) %>%
    # Coalesce to prioritize exact matches
    dplyr::mutate(
      matched_accession = coalesce(accession, accession_base),
      final_submission_url = coalesce(submission_url, submission_url_base),
      final_zip_url = coalesce(zip_url, zip_url_base),
      final_summary_url = coalesce(summary_url, summary_url_base)
    ) %>%
    dplyr::select(
      data, 
      species, 
      matched_accession,
      submission_url = final_submission_url,
      zip_url = final_zip_url,
      summary_url = final_summary_url
    ) %>% 
    dplyr::filter(!is.na(submission_url))

# Export your accessions to a text file
write.csv(
  phaster_complete_sequences_df %>%
    dplyr::pull(data),
    file = "prophage/phaster_accessions.txt",
    quote = F,
    row.names = F,
    col.names = FALSE)

```

``` {bash, eval = FALSE}

cd prophage
./get_phaster_ncbi_info.sh

```

``` {r Parse genome lengths to identify complete sequences, fig.width = 8, fig.height = 10}

seq_length_df <-
  read.table("./prophage/phaster_metadata.tsv",
             sep = "\t",
             comment.char = "",
             header = TRUE) %>% 
    dplyr::left_join(phaster_complete_sequences_df,
                     by = c("Accession"="data")) %>% 
    dplyr::group_by(species,BioSampleAccession) %>% 
    dplyr::mutate(Total_length = sum(SequenceLength)) %>% 
    dplyr::ungroup() %>% 
    dplyr::left_join(
      enframe(lower_length_threshold) %>%
            tidyr::unnest(value) %>% 
            dplyr::rename(Lower = value) %>% 
        dplyr::left_join(
        enframe(upper_length_threshold) %>%
            tidyr::unnest(value) %>% 
            dplyr::rename(Upper = value)
        ) %>%
        dplyr::left_join(enframe(species_name) %>%
                           tidyr::unnest(value) %>% 
                           dplyr::rename(species = value),
                             by = "name") %>%
        dplyr::left_join(enframe(display_name) %>%
                           tidyr::unnest(value) %>% 
                           dplyr::rename(display = value),
                             by = "name")
      ) %>% 
      dplyr::filter(!is.na(name))

# Plot genome lengths and boundaries
ggplot(seq_length_df,
       aes(x = Total_length)) +
  geom_histogram(colour = "blue", fill = "blue") +
  geom_vline(data = seq_length_df %>%
               dplyr::select(display, Upper) %>% 
               dplyr::distinct(),
             aes(xintercept = Upper),
             colour = "red", linetype = 2) +
  geom_vline(data = seq_length_df %>%
               dplyr::select(display, Lower) %>% 
               dplyr::distinct(),
             aes(xintercept = Lower),
             colour = "red", linetype = 2) +
  facet_wrap(~display,
             scales = 'free_y',
             labeller = label_parsed) +
  ylab("Count") +
  xlab("Total genome length (bp)") +
  theme_bw() +
  theme(strip.background = element_blank(),
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

```

``` {r Filter datasets by length}

filtered_genome_df <-
  seq_length_df %>% 
    dplyr::filter(Total_length >= Lower & Total_length <= Upper)

write.table(filtered_genome_df %>% 
            dplyr::mutate(summary_url = gsub("/submissions/","/jobs/",summary_url)) %>% 
            dplyr::select(summary_url),
          file = "./phaster_outputs/phaster_urls.list",
          row.names = FALSE,
          col.names = FALSE,
          quote = FALSE)

```

Each dataset was then retrieved from the database:

``` {bash, eval = FALSE}

cd phaster_outputs
./download_phaster.sh

```

``` {r Processing phaster summary.txt, eval = FALSE}

parse_phaster_summary <- function(file, accession) {
  # If file missing or empty, return nothing
  if (!file.exists(file)) return(NULL)
  lines <- readr::read_lines(file)
  if (length(lines) == 0) return(NULL)

  # Find header line
  header_idx <- which(str_detect(lines, "^\\s*REGION\\s+REGION_LENGTH"))
  if (length(header_idx) == 0) return(NULL)

  # Find dashed separator after header
  sep_idx <- header_idx + 1
  while (sep_idx <= length(lines) && !str_detect(lines[sep_idx], "-{5,}")) {
    sep_idx <- sep_idx + 1
  }
  if (sep_idx > length(lines)) return(NULL)

  data_start <- sep_idx + 1

  # Data lines
  data_lines <- lines[data_start:length(lines)]
  data_lines <- data_lines[data_lines != ""]
  if (length(data_lines) == 0) return(NULL)

  # --- line parser using whitespace ---
  parse_line <- function(x) {
    parts <- x %>%
      str_squish() %>%
      str_split("\\s+", simplify = TRUE) %>%
      as.character()

    if (length(parts) < 16) return(NULL)

    region        <- parts[1]
    region_length <- parts[2]
    completeness  <- parts[3]

    rest <- parts[-(1:3)]
    if (length(rest) == 0) return(NULL)

    # REGION_POSITION looks like "start-end"
    pos_idx <- which(str_detect(rest, "^[0-9]+-[0-9]+$"))[1]
    if (is.na(pos_idx)) return(NULL)

    # SPECIFIC_KEYWORD = everything before REGION_POSITION
    specific_keyword <-
      if (pos_idx == 1) NA_character_
      else paste(rest[1:(pos_idx - 1)], collapse = " ")

    tail_tokens <- rest[pos_idx:length(rest)]
    if (length(tail_tokens) < 13) return(NULL)

    tibble(
      REGION                    = region,
      REGION_LENGTH             = region_length,
      COMPLETENESS_SCORE        = completeness,
      SPECIFIC_KEYWORD          = specific_keyword,
      REGION_POSITION           = tail_tokens[1],
      TRNA_NUM                  = tail_tokens[2],
      TOTAL_PROTEIN_NUM         = tail_tokens[3],
      PHAGE_HIT_PROTEIN_NUM     = tail_tokens[4],
      HYPOTHETICAL_PROTEIN_NUM  = tail_tokens[5],
      PHAGE_HYPO_PERCENT        = tail_tokens[6],
      BACTERIAL_PROTEIN_NUM     = tail_tokens[7],
      ATT_SITE_SHOWUP           = tail_tokens[8],
      PHAGE_SPECIES_NUM         = tail_tokens[9],
      MOST_COMMON_PHAGE_NAME    = tail_tokens[10],
      MOST_COMMON_PHAGE_NUM     = tail_tokens[11],
      MOST_COMMON_PHAGE_PERCENT = tail_tokens[12],
      GC_PERCENTAGE             = tail_tokens[13]
    )
  }

  df <- map_dfr(data_lines, parse_line)
  if (!("REGION" %in% names(df)) || nrow(df) == 0) return(NULL)

  # ---------- clean types, keep all regions ----------
  clean_df <- df %>%
    filter(str_detect(REGION, "^\\d+$")) %>%
    mutate(
      REGION              = as.integer(REGION),
      REGION_LENGTH_Kb    = as.numeric(str_remove(REGION_LENGTH, "Kb")),
      COMPLETENESS        = str_extract(COMPLETENESS_SCORE, "^[^()]+") %>% str_trim(),
      SCORE               = str_extract(COMPLETENESS_SCORE, "(?<=\\()\\d+(?=\\))") %>% as.integer(),
      TOTAL_PROTEIN_NUM   = as.integer(TOTAL_PROTEIN_NUM),
      TRNA_NUM            = as.integer(TRNA_NUM),
      PHAGE_HIT_PROTEIN_NUM    = as.integer(PHAGE_HIT_PROTEIN_NUM),
      HYPOTHETICAL_PROTEIN_NUM = as.integer(HYPOTHETICAL_PROTEIN_NUM),
      PHAGE_HYPO_PERCENT       = as.numeric(str_remove(PHAGE_HYPO_PERCENT, "%")),
      BACTERIAL_PROTEIN_NUM    = as.integer(BACTERIAL_PROTEIN_NUM),
      PHAGE_SPECIES_NUM        = as.integer(PHAGE_SPECIES_NUM),
      MOST_COMMON_PHAGE_NUM    = as.integer(MOST_COMMON_PHAGE_NUM),
      MOST_COMMON_PHAGE_PERCENT = as.numeric(str_remove(MOST_COMMON_PHAGE_PERCENT, "%")),
      GC_PERCENTAGE            = as.numeric(str_remove(GC_PERCENTAGE, "%"))
    )

  if (nrow(clean_df) == 0) return(NULL)

  # ---------- keywords → long ----------
  keywords_long <- clean_df %>%
    mutate(SPECIFIC_KEYWORD = coalesce(SPECIFIC_KEYWORD, "")) %>%
    separate_rows(SPECIFIC_KEYWORD, sep = ",") %>%
    mutate(
      SPECIFIC_KEYWORD = str_trim(SPECIFIC_KEYWORD),
      SPECIFIC_KEYWORD = na_if(SPECIFIC_KEYWORD, "")
    ) %>%
    filter(!is.na(SPECIFIC_KEYWORD))

  # No keywords at all in this file: return regions without keyword columns
  if (nrow(keywords_long) == 0) {
    out <- clean_df %>%
      transmute(
        accession,
        REGION,
        REGION_LENGTH_Kb,
        TOTAL_PROTEIN_NUM,
        COMPLETENESS,
        SCORE,
        MOST_COMMON_PHAGE_NAME
      )
    return(out)
  }

  # ---------- keywords → wide, values_fill = 0 ----------
  keyword_matrix <- keywords_long %>%
    dplyr::mutate(value = 1L) %>%
    dplyr::select(REGION, SPECIFIC_KEYWORD, value) %>%
    dplyr::distinct() %>%
    tidyr::pivot_wider(
      names_from  = SPECIFIC_KEYWORD,
      values_from = value,
      values_fill = 0   # ensures missing combos get 0
    )

  # ---------- final output ----------
  final_df <- clean_df %>%
    dplyr::select(
      REGION,
      REGION_LENGTH_Kb,
      TOTAL_PROTEIN_NUM,
      COMPLETENESS,
      SCORE,
      MOST_COMMON_PHAGE_NAME
    ) %>%
    dplyr::left_join(keyword_matrix, by = "REGION") %>%
    dplyr::mutate(accession = accession,.before = 1)

  # Ensure *all* keyword columns are 0/1 with no NA
  # (i.e. any NA left from the join are turned into 0)
  non_keyword_cols <- c(
    "accession",
    "REGION",
    "REGION_LENGTH_Kb",
    "TOTAL_PROTEIN_NUM",
    "COMPLETENESS",
    "SCORE",
    "MOST_COMMON_PHAGE_NAME"
  )

  keyword_cols <- setdiff(names(final_df), non_keyword_cols)

  if (length(keyword_cols) > 0) {
    final_df <- final_df %>%
      dplyr::mutate(across(all_of(keyword_cols),
                    ~ replace_na(as.numeric(.x), 0)))
  }

  final_df
}

all_results <- purrr::map_dfr(
  filtered_genome_df$Accession,
  ~parse_phaster_summary(
    file      = paste0("./phaster_outputs/",.x, ".txt"),
    accession =.x
  )
) %>%
  { 
    non_keyword_cols <- c(
      "accession",
      "REGION",
      "REGION_LENGTH_Kb",
      "TOTAL_PROTEIN_NUM",
      "COMPLETENESS",
      "SCORE",
      "MOST_COMMON_PHAGE_NAME"
    )
    keyword_cols <- setdiff(names(.), non_keyword_cols)

    dplyr::mutate(.,
      across(
        all_of(keyword_cols),
        ~ replace_na(as.numeric(.x), 0)
      )
    )
  }

# Store summarised results for more efficient reanalysis
write_csv(all_results,
          file = "./phaster_outputs/phaster_analysis_summary.csv"
          )

```

``` {r Plot phaster outputs, fig.width = 8}

# Read phaster summary
all_results <-
  read_csv("./phaster_outputs/phaster_analysis_summary.csv")

# Plot output
ggplot(all_results %>%
         dplyr::mutate(
           Lytic = dplyr::if_else(lysis == 0,
                                  "No lysis genes detected",
                                  "Lysis genes present")
         ) %>%
         dplyr::mutate(
           COMPLETENESS = factor(str_to_sentence(COMPLETENESS),
                                 levels = c(
                                   "Intact",
                                   "Questionable",
                                   "Incomplete"
                                 )
           )
         ),
       aes(x = REGION_LENGTH_Kb)) +
  geom_histogram(colour = "blue", fill = "blue") +
  scale_x_log10() +
  ylab("Count") +
  xlab("Region length (kb)") +
  geom_vline(xintercept = 30, colour = "red", linetype = 2) +
  facet_grid(Lytic~COMPLETENESS) +
  theme_bw() +
  theme(strip.background = element_blank())

```

``` {r Tabulate number of samples per species}

# Get summaries of MLST data
st_counts_df <-
  read_csv(file = "./MLST_data/st_counts.csv")

# Combine summaries
sample_summary_df <-
  tibble::enframe(table(complete_sequences_df$species)) %>% 
    dplyr::rename(`Complete replicons` = value) %>% 
    dplyr::left_join(
      tibble::enframe(table(phaster_complete_sequences_df$species)) %>% 
        dplyr::rename(`PHASTER entries` = value),
      by = "name"
    ) %>% 
    dplyr::left_join(
      st_counts_df,
      by = "name"
    ) %>% 
    dplyr::left_join(
      as.data.frame(table(mismatches_df$Species)) %>% 
        dplyr::rename(
          name = Var1,
          `Single locus variants` = Freq
        ) %>% 
        dplyr::left_join(enframe(species_name) %>% 
                           tidyr::unnest(value),
                         by = "name") %>% 
        dplyr::mutate(name = value) %>% 
        dplyr::select(-value),
      by = "name"
    ) %>% 
    dplyr::left_join(
      filtered_genome_df %>% 
        dplyr::group_by(species) %>% 
        dplyr::summarise(`Full-length genomes` = n_distinct(BioSampleAccession)) %>% 
        dplyr::ungroup(),
      by = c("name"="species")
    ) %>% 
    dplyr::rename(Species = name) %>% 
    dplyr::select(Species,`Sequence types`,`Single locus variants`,`Complete replicons`,`PHASTER entries`,`Full-length genomes`)

# Generate summary table
sample_summary_df %>%
  kable() %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)

```


``` {r Count prophage per genome, fig.height = 8, fig.width = 8}

# Filter prophage and count per genome
filtered_prophage_count_df <-
  all_results %>% 
    dplyr::filter(REGION_LENGTH_Kb >= 30 & REGION_LENGTH_Kb <= 150 & COMPLETENESS == "intact") %>% 
    dplyr::group_by(accession) %>% 
    dplyr::summarise(phage_count = n()) %>% 
    dplyr::ungroup()

# Combine phage counts with species information
filtered_prophage_per_sample_df <-
  filtered_genome_df %>% 
    # Set consistent name
    dplyr::group_by(BioSampleAccession) %>% 
    dplyr::mutate(Name = dplyr::if_else(SequenceLength==max(SequenceLength),BioSampleName,NA_character_)) %>% 
    tidyr::fill(Name, .direction = "updown") %>% 
    dplyr::ungroup() %>% 
    # Process prophage data
    dplyr::left_join(filtered_prophage_count_df, by = c("Accession" = "accession")) %>% 
    dplyr::mutate(phage_count = dplyr::if_else(is.na(phage_count),0,phage_count)) %>% 
    dplyr::group_by(BioSampleAccession,Name,species) %>% 
    dplyr::summarise(total_phage_count = sum(phage_count)) %>% 
    dplyr::ungroup() %>% 
    dplyr::group_by(species) %>%
    dplyr::mutate(
      total_obs = n()
    ) %>%
    dplyr::ungroup() %>% 
    dplyr::group_by(species,total_phage_count) %>%
    dplyr::mutate(
      obs_prob = n()/total_obs
    ) %>%
    dplyr::ungroup() %>% 
    dplyr::left_join(
      enframe(species_name) %>% 
        tidyr::unnest(value) %>% 
        dplyr::rename(species = value),
      by = "species"
    ) %>% 
    dplyr::left_join(
      enframe(display_name) %>% 
        tidyr::unnest(value) %>% 
        dplyr::rename(display = value),
      by = "name"
    )

# Fit statistical distributions to prophage count data for each species
fits_per_species <-
  filtered_prophage_per_sample_df %>%
  dplyr::group_by(species) %>%
  dplyr::summarise(
    data = list(total_phage_count),.groups = "drop"
  ) %>%
  dplyr::mutate(
    # Fit Poisson
    fit_pois = map(data, ~ suppressWarnings(fitdist(unlist(.x), "pois", method = "mle")))
  )

# Extract fit information
pois_params <- fits_per_species %>%
  dplyr::transmute(
    species,
    lambda = map_dbl(fit_pois, ~.x$estimate[["lambda"]])
  )

# Extract confidence intervals
pois_ci <-
  fits_per_species %>%
    dplyr::mutate(
      # extract point estimate
      lambda_hat = map_dbl(fit_pois, ~.x$estimate[["lambda"]]),
      # extract CI as a matrix, then convert to vector
      ci_mat = map(fit_pois, ~ confint(.x))  # default 95% Wald CI
    ) %>%
    dplyr::mutate(
      lambda_lwr = map_dbl(ci_mat, ~.x["lambda", "2.5 %"]),
      lambda_upr = map_dbl(ci_mat, ~.x["lambda", "97.5 %"])
    ) %>%
    dplyr::mutate(lambda_lwr = dplyr::if_else(lambda_lwr < 0, 0.0, lambda_lwr)) %>% 
    dplyr::select(species, lambda_hat, lambda_lwr, lambda_upr)

# Calculate probabilities
count_grid <- tibble(total_phage_count = 0:max(filtered_prophage_per_sample_df$total_phage_count))
fitted_df <-
  pois_params %>% 
    dplyr::select(species, lambda) %>%
    tidyr::crossing(count_grid) %>%
    dplyr::mutate(
        dist = "Poisson",
        fitted_prob = dpois(total_phage_count, lambda)
    ) %>% 
    dplyr::left_join(
      enframe(species_name) %>% 
        tidyr::unnest(value) %>% 
        dplyr::rename(species = value),
      by = "species"
    ) %>% 
    dplyr::left_join(
      enframe(display_name) %>% 
        tidyr::unnest(value) %>% 
        dplyr::rename(display = value),
      by = "name"
    )

# Plot fitted distributions against observed prophage counts
ggplot() +
  # Observed probabilities as bars
  geom_col(
    data = filtered_prophage_per_sample_df %>% 
            dplyr::select(species,total_phage_count,obs_prob,display) %>% 
            dplyr::distinct(),
    aes(x = total_phage_count, y = obs_prob),
    fill = "blue",
    color = "blue",
    width = 0.8
  ) +
  # Fitted distributions as lines
  geom_line(
    data = fitted_df,
    aes(x = total_phage_count,
        y = fitted_prob),
    size = 0.5,
    linetype = 2,
    colour = "red"
  ) +
  facet_wrap(~display,
             scales = "free_y",
             labeller = label_parsed) +
  labs(
    x = "Total phage count per sample",
    y = "Proportion",
    color = "Distribution"
  ) +
  theme_bw() +
  theme(strip.background = element_blank(),
        legend.position = "bottom")

```

Two groups of bacteria exhibited a substantial deviation from the unimodal distribution expected of a homogeneously-behaving population:

- *Enterobacter* spp.: there are many isolates (of both *E. hormaechei* and *E. cloacae*) that lack any prophage, despite a peak at 4 prophage per isolate (isolates of *E. hormaechei*)

- *Proteus* spp.: there are many isolates (of both *P. mirabilis* and *P. terrae*) that lack any prophage, despite a peak at 2 prophage per isolate (isolates of *P. mirabilis* and *P. terrae*)

These were excluded from the subsequent comparison, to ensure the MLST and prophage data both corresponded to homogeneously behaving populations of isolates.

# Compare prophage density and recombination rate

Model the relationship between the density of prophage in genomes and rates of core genome recombination:

``` {r Compare prophage density with recombination rate, fig.height = 8, fig.width = 8}

# Excluded species
excluded_species <- c("Esp","Psp")

# Combine prophage density with homologous recombination
prophage_rec_df <-
  rec_prob_df %>% 
    dplyr::rename(`Recombination probability` = prob) %>% 
    dplyr::left_join(tibble::enframe(species_name) %>%
                       tidyr::unnest(value) %>% 
                       dplyr::rename(
                         Species = name,
                         species_name = value
                       ),
                     by = "Species") %>% 
    dplyr::left_join(tibble::enframe(display_name) %>%
                       tidyr::unnest(value) %>% 
                       dplyr::rename(
                         Species = name,
                         display_name = value
                       ),
                     by = "Species") %>% 
    dplyr::ungroup() %>% 
    dplyr::left_join(filtered_prophage_per_sample_df %>% 
                      dplyr::group_by(species) %>% 
                      dplyr::summarise(mean_prophage = mean(total_phage_count)) %>% 
                      dplyr::ungroup(),
                     by = c("species_name" = "species")) %>% 
    dplyr::left_join(pois_ci,
                     by = c("species_name" = "species")) %>% 
    # Add transformable status
    dplyr::left_join(enframe(transformable) %>%
                      tidyr::unnest(value) %>% 
                      dplyr::rename(Transformability = value),
                     by = c("Species" = "name")) %>% 
    # Add ecological status
    dplyr::left_join(enframe(primary_niche) %>%
                      tidyr::unnest(value) %>% 
                      dplyr::rename(Niche = value),
                     by = c("Species" = "name")) %>% 
    dplyr::filter(!(Species %in% excluded_species) & !is.na(species_name))

# Fit model of recombination rate and prophage density
rec_prophage_model <-
  glmmTMB::glmmTMB(
    `Recombination probability` ~ mean_prophage,
    data = prophage_rec_df,
    ziformula = ~0,
    dispformula = ~1,
    family = beta_family(link="logit")
  )

performance::check_model(rec_prophage_model)

```

Plot the data and model output together with bacterial species information:

``` {r Plot rec phage model output, fig.width = 8, fig.height = 6}

# Create a grid of mean_prophage values over the observed range
newdata_rec <- prophage_rec_df %>%
  summarise(
    min_mean = min(mean_prophage, na.rm = TRUE),
    max_mean = max(mean_prophage, na.rm = TRUE)
  ) %>%
  { tibble(mean_prophage = seq(.$min_mean,.$max_mean, length.out = 200)) }

emm_rec <- emmeans(
  rec_prophage_model,
  specs = ~ mean_prophage,
  at    = list(mean_prophage = newdata_rec$mean_prophage),
  type  = "response"   # back-transform from logit to probability
)

rec_prophage_model_predictions <- as_tibble(emm_rec) %>%
  # emmeans returns columns: mean_prophage, response, SE, lower.CL, upper.CL
  dplyr::rename(
    prob_fit = response,
    prob_se  = SE,
    prob_lwr = asymp.LCL,
    prob_upr = asymp.UCL
  )

###

# Extract p value
mod_sum<-summary(rec_prophage_model)
p_val  <- mod_sum$coefficients$cond["mean_prophage", "Pr(>|z|)"]
p_label <- paste0("italic(p) == ", signif(p_val, 3))

# Define palettes
niche_palette <- c(
  "Environmental" = 8,
  "Gastrointestinal tract" = 16,
  "Respiratory tract" = 1,
  "Multiple" = 5
)

transformability_palette <- c(
  "Transformable" = "cornflowerblue",
  "Non-transformable" = "purple"
)

# Generate plot combining prophage and recombination data
(comparison_plot <-
  ggplot(prophage_rec_df,
         aes(y = `Recombination probability`,
             x = mean_prophage,
             colour = Transformability,
             shape = Niche,
             label = display_name)) +
    geom_point() +
    geom_errorbar(aes(ymin = asymp.LCL,
                      ymax = asymp.UCL),
                  alpha = 0.5) +
    geom_errorbarh(aes(xmin = lambda_lwr,
                       xmax = lambda_upr),
                  alpha = 0.5) +
    # fitted model line
    geom_line(
      data = rec_prophage_model_predictions,
      aes(x = mean_prophage, y = prob_fit),
      inherit.aes = FALSE,
      color = "black",
      linewidth = 1,
      alpha = 0.75
    ) +
    # confidence band for the fit
    geom_ribbon(
      data = rec_prophage_model_predictions,
      aes(x = mean_prophage, ymin = prob_lwr, ymax = prob_upr),
      inherit.aes = FALSE,
      alpha = 0.2,
      fill = "black"
    ) +
    ylab("Probability of diversification through recombination") +
    xlab("Expected number of prophage in an isolate genome") +
    scale_shape_manual(values = niche_palette) +
    scale_colour_manual(values = transformability_palette,
                        aesthetics = c("colour","fill")) +
    ggrepel::geom_label_repel(size = 3,
                             force_pull = 0.5,
                             parse = TRUE) +
    guides(
      colour = guide_legend(title = "Transformability",
                            nrow = 1),
      shape  = guide_legend(title = "Niche",
                            nrow = 1)
    ) +
    theme_bw() +
    theme(legend.position = "bottom",
          legend.box = "vertical") +
    annotate(
      "text",
      x = 4,
      y = 0.7,
      label = p_label,
      hjust = 1.1, vjust = 1.5,
      parse = TRUE
    )
)

```

# Experimental data

## Plot qPCR data

``` {r Plot qPCR data, fig.width = 8, fig.height = 6}

competence_palette <- c(
  "Spontaneous competence" = "darkolivegreen3",
  "Induced competence" = "seagreen3",
  "Control" = "grey"
)

qpcr_df <-
  read_csv("./experimental_data/qPCR_data.csv") %>% 
    dplyr::mutate(`Experimental conditions` =
                    dplyr::case_when(
                      mutants == "NO_CSP" ~ "Control",
                      mutants == "CSP_DnaseI" ~ "Control",
                      mutants == "CSP" ~ "Control",
                      grepl("_CSP",mutants) ~ "Induced competence",
                      TRUE ~ "Spontaneous competence"
                    )
    ) %>% 
    dplyr::mutate(DNA =
                    dplyr::case_when(
                      mutants == "NO_CSP" ~ "No CSP control",
                      mutants == "CSP_DnaseI" ~ "CSP and DNase",
                      mutants == "CSP" ~ "No DNA control",
                      grepl("E99",mutants) ~ "1% pneumococcal",
                      grepl("E90",mutants) ~ "10% pneumococcal",
                      grepl("E50",mutants) ~ "50% pneumococcal",
                      grepl("E10",mutants) ~ "90% pneumococcal",
                      grepl("E1",mutants) ~ "99% pneumococcal"
                    )
    ) %>% 
  dplyr::mutate(
    Timepoint = 
      dplyr::case_when(
        conditions == "2hour" ~ "'2h post-CSP'",
        conditions == "4hour" ~ "'4h post-CSP'"
      )
  ) %>% 
  tidyr::pivot_longer(cols = c(Phage_circular,PRCI_circular),
                      names_to = "MGE",
                      values_to = "qPCR") %>% 
  dplyr::mutate(
    RMV8_MGE = 
      factor(
        dplyr::case_when(
          MGE == "Phage_circular" ~ "phi*'RMV8'",
          MGE == "PRCI_circular" ~ "'PRCI'[italic(malA)]"
        ),
        levels = c("phi*'RMV8'","'PRCI'[italic(malA)]")
      )
  ) %>% 
  dplyr::filter(!is.na(qPCR)) %>% 
  dplyr::mutate(`Biological replicate` = as.character(Replicate))

(qpcr_plot <-
  ggplot(qpcr_df,
         aes(x = DNA,
             y = qPCR,
             colour = `Experimental conditions`,
             group = interaction(DNA,`Experimental conditions`),
             shape = `Biological replicate`)) +
    geom_violin(draw_quantiles = c(0.5),
                position = position_dodge(preserve = "single"),
                scale = "width",) +
    geom_point(position = position_jitterdodge()) +
    scale_colour_manual(values = competence_palette,
                        aesthetics = c("fill","colour")) +
    scale_y_continuous(trans = "log10") +
    ylab(expression(atop(italic(attP)~'sites in','culture supernatant'))) +
    facet_grid(RMV8_MGE~Timepoint,
               labeller = label_parsed) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
          strip.background = element_blank())
)

```

``` {r Focus on change in spontaneous deletion}

spontaneous_start_end_df <-
  dplyr::bind_rows(
     qpcr_df %>%
      dplyr::filter(
        `Experimental conditions` == "Spontaneous competence" & 
        MGE == "Phage_circular" & 
        conditions == "4hour" & 
        DNA %in% c("1% pneumococcal","99% pneumococcal")
      ),
    dplyr::bind_rows(
      read_csv("./experimental_data/additional_qPCR_data.csv"),
      read_csv("./experimental_data/Spontaneous_extra_replicates.csv")
    ) %>% 
      dplyr::filter(conditions=="4hour") %>% 
      dplyr::mutate(`Experimental conditions` = "Spontaneous competence") %>% 
      dplyr::mutate(DNA =
                      dplyr::case_when(
                        grepl("E99",mutants) ~ "1% pneumococcal",
                        grepl("E1",mutants) ~ "99% pneumococcal"
                      )
      ) %>% 
      tidyr::pivot_longer(cols = c(Phage_circular,PRCI_circular),
                          names_to = "MGE",
                          values_to = "qPCR") %>% 
      dplyr::filter(MGE == "Phage_circular") %>% 
      dplyr::mutate(`Biological replicate` = as.character(Replicate)),
    read_csv("./experimental_data/extra_replicate_spontaneous_qPCR.csv") %>% 
      dplyr::filter(conditions=="4hour") %>% 
      dplyr::mutate(DNA =
                      dplyr::case_when(
                        grepl("E99",mutants) ~ "1% pneumococcal",
                        grepl("E1",mutants) ~ "99% pneumococcal"
                      )
      ) %>% 
      tidyr::pivot_longer(
        cols = c(Phage_circular,PRCI_circular),
        names_to = "MGE",
        values_to = "qPCR") %>% 
      dplyr::filter(MGE == "Phage_circular") %>% 
      dplyr::mutate(`Biological replicate` = as.character(Replicate))
  ) %>% 
    dplyr::group_by(DNA,`Biological replicate`) %>% 
    dplyr::mutate(Mean_est = mean(qPCR)) %>% 
    dplyr::ungroup() %>% 
    dplyr::group_by(DNA) %>% 
    dplyr::mutate(Baseline =
                    dplyr::if_else(
                      DNA == "1% pneumococcal",
                      Mean_est,
                      NA_real_
                    )
                  ) %>% 
    dplyr::ungroup() %>% 
    dplyr::group_by(`Biological replicate`) %>% 
    tidyr::fill(Baseline, .direction = "updown") %>% 
    dplyr::mutate(Change = qPCR/Baseline) %>% 
    dplyr::ungroup() %>% 
    dplyr::filter(DNA == "99% pneumococcal")

# Plot fold changes
(fold_change_plot <-
  ggplot(spontaneous_start_end_df,
         aes(x = `Biological replicate`,
             y = Change)) +
    geom_violin(draw_quantiles = c(0.5),
                scale = "width",
                colour = competence_palette[["Spontaneous competence"]]) +
    geom_point(position = position_jitter(width = 0.15),
               colour = competence_palette[["Spontaneous competence"]]) +
    scale_y_continuous(trans = "log10") +
    geom_vline(xintercept = 3.5, linetype = 2) +
    ylab(expression("Fold change in"~italic(attP)~"concentration")) +
    theme_bw() +
    theme(axis.text.x = element_blank(),
          strip.background = element_rect(fill = NA))
)

```

Plot transformation data:

``` {r Plot transformation data}

transformation_df <-
  dplyr::bind_rows(
    read_csv("./experimental_data/transformation_rates.csv"),
    read_csv("./experimental_data/deletions_with_CSP_three_replicates.csv") %>% 
      dplyr::filter(conditions == "Phage_Kan") %>% 
      dplyr::select(-conditions) %>% 
      dplyr::mutate(mutants = gsub("_","+",donor)) %>% 
      dplyr::rename(deletion_rates = Transformation_efficiency)
  ) %>% 
  dplyr::mutate(DNA =
                    dplyr::case_when(
                      mutants == "Control" ~ "No CSP control",
                      mutants == "CSP_DnaseI" ~ "CSP and DNase",
                      mutants == "CSP" ~ "No DNA control",
                      grepl("E99",mutants) ~ "1% pneumococcal",
                      grepl("E90",mutants) ~ "10% pneumococcal",
                      grepl("E50",mutants) ~ "50% pneumococcal",
                      grepl("E10",mutants) ~ "90% pneumococcal",
                      grepl("E1",mutants) ~ "99% pneumococcal"
                    )
    ) %>% 
    dplyr::mutate(`Experimental conditions` =
                    dplyr::case_when(
                      mutants == "Control" ~ "Control",
                      mutants == "CSP_DnaseI" ~ "Control",
                      mutants == "CSP" ~ "Control",
                      grepl("CSP",mutants) ~ "Induced competence",
                      TRUE ~ "Spontaneous competence"
                    )
    )

# Plot transformation data
(transformation_plot <-
  ggplot(transformation_df,
         aes(x = DNA,
             y = deletion_rates + 1e-8,
             colour = `Experimental conditions`,
             group = interaction(mutants,`Experimental conditions`))) +
    geom_violin(draw_quantiles = c(0.5),
                position = position_dodge(preserve = "single"),
                scale = "width") +
    geom_point(position = position_jitterdodge()) +
    ylab(expression(atop("Transformation efficiency","(transformants per cfu)")))+
    scale_colour_manual(values = competence_palette) +
    scale_y_continuous(trans = "log10") +
    geom_hline(yintercept = 1e-8, linetype = 3) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
          strip.background = element_rect(fill = NA))
)

```

Combine experimental data into a single figure:

``` {r Combined experimental data figure, fig.width = 8, fig.height = 8}

legend_plot <-
  cowplot::get_legend(qpcr_plot)

bottom_right_plot <-
  cowplot::plot_grid(plotlist = list(fold_change_plot +
                                       theme(legend.position = "none"),
                                     legend_plot),
                                     nrow = 1,
                                     labels = c("C","")
                     )

bottom_plots <-
  cowplot::plot_grid(plotlist = list(transformation_plot +
                                       theme(legend.position = "none"),
                                     bottom_right_plot),
                     labels = c("B",""),
                     nrow = 1,
                     rel_widths = c(0.5,0.5))

(exp_plot <-
  cowplot::plot_grid(plotlist = list(qpcr_plot +
                                       theme(legend.position = "none"),
                                     bottom_plots),
                     labels = c("A",""),
                     nrow = 2)
)

```

Now plot the control DNase data:

``` {r Plot DNase control data, fig.width = 6, fig.height = 4}

dnase_control_df <-
  read_csv("./experimental_data/DnaseI_control.csv") %>% 
    tidyr::pivot_longer(
      cols = c(Ratio_PRCI,Ratio_Prophage),
      names_to = "MGE",
      values_to = "qPCR"
    ) %>% 
    dplyr::mutate(
      Conditions =
        dplyr::if_else(
          grepl("No_DnaseI",conditions),
          "No DNase",
          "DNase"
          )
      ) %>% 
    dplyr::mutate(
      Ratio = 
        dplyr::case_when(
          MGE == "Ratio_PRCI" ~ "PRCI[italic(malA)]~'to'~italic(rpoA)",
          MGE == "Ratio_Prophage" ~ "phi*'RMV8'~'to'~italic(rpoA)"
        )
    ) %>% 
  dplyr::rename(`Biological replicate` = replicate) %>% 
  dplyr::mutate(
    Experiment = factor(
      dplyr::case_when(
        mutants == "control" ~ "No CSP control",
        mutants == "CSP_induced" ~ "4h post-CSP"
      ),
      levels = c("No CSP control","4h post-CSP")
    )
  )

# Define DNase palette
dnase_palette <- c(
  "DNase" = "red",
  "No DNase" = "grey"
)

# Plot DNase control
ggplot(dnase_control_df,
       aes(x = Experiment,
           y = qPCR,
           shape = `Biological replicate`,
           group = interaction(mutants,Conditions),
           colour = Conditions)) +
  geom_violin(position = position_dodge(preserve = "single"),
              draw_quantiles = c(0.5)) +
  geom_point(position = position_jitterdodge(jitter.width = 0.15)) +
  scale_colour_manual(values = dnase_palette) +
  facet_grid(~Ratio,
               labeller = label_parsed) +
  ylab(expression('Relative DNA amount ('*2^{-Delta*Delta*'C'[t]}*')')) +
  scale_y_log10() +
  theme_bw() +
  theme(strip.background = element_blank())

```
